---
title: |
  | Assignment 3
  | Visualisation in R
author: |
  | Your Name Goes Here
output: word_document
---

##Note:
Using the mpg data set (base in R) we will explore ggplot2 and what the added plotting features can do.
Please remember to comment all your code and describe your findings in detail.
This assignment is intended for you to work alone on, please do not use out side resources like code we do not cover in class or other students. Everything in this assignment is covered in week 3 and prior weeks. If you have any questions please reach out via email.

##Please Load packages and data:
```{r}

library(ggplot2) # loads the ggplot2 package
library(dplyr)
# loads the dplyr package 
data <- mpg 
# loads the data set mpg  and assigns to data variable
data
# prints the data set
```
1) Explore the data set, what are the variables?
```{r}
# Load required library
library(ggplot2)
data <- mpg  
# Print the column names (variables) in the dataset
print(names(data))  # Displays the names of all columns (variables)
# Display the structure of the dataset (data types, sample values)
str(data)
# Display the first few rows of the dataset
head(data)
# Summary statistics of the dataset
summary(data)

```

2) Create a bar chart of 'cyl' count for the data set.
```{r}
ggplot(data, aes(x = factor(cyl))) + 
  #  dataset as data and aes mapping to "cyl" as x axis
  geom_bar() + # plots the bar graph
  labs(x = "Number of Cylinders", y = "car count", title = "car count on no of cylinder it has") 
# main title is added and the description for X and Y axis
```

3) Plot two variables (cty vs hwy) fuel economy as points, do we see a pattern?
```{r}
ggplot(data, aes(x = cty, y = hwy)) + 
  # aes() mapping  "cty" for x-axis and "hwy" for y- axis 
  geom_point() + # creates scatter plot 
  labs(x = "City MPG", # x-axis label
       y = "Highway MPG", # y-axis label
       title = "City MPG  vs. Highway MPG") # title for chart

# As city fuel economy increases, highway fuel economy also increases.
#the cars that provides good mileage in the city also gives good mileage on the highway.
#Some variation may exist due to vehicle types and driving conditions.
```

4) Build a Box and wisker plot of the car 'class' and cty fuel economy 
```{r}
ggplot(data, aes(x = class, y = cty)) +
  # mapping  "class" on x and "cty" on y axes
  geom_boxplot() + 
  # boxplot is plotted
  labs(x = "Car Class", # x-axis label
       y = "City MPG", # y-axis label
       title = "City MPG by Car Class") # title for graph
 
#Compact and subcompact cars tend to have higher city MPG.
# SUVs and pickups have lower city MPG, showing less fuel efficiency

```

5) Lets plot cty vs hwy again but now include the class "cyl" and color code it by number of cyl
```{r}
ggplot(data, aes(x = cty, y = hwy, color = factor(cyl))) + 
  # plotting cty for x axis, hwy for y axis and color as cyl)
  geom_point() + # adds points
  labs(x = "City MPG", # x-axis label
       y = "Highway MPG", # y-axis label
       title = "City vs. Highway MPG by Number of Cylinders") # title of the chart
# the color variation by the number of cylinders explains that cars with fewer normally have higher efficiency in both city and highway driving situations.
```

6) Let's build on chart 5, using 'facet_wrap()' lets compare 1999 to 2008.
```{r}
ggplot(data, aes(x = cty, y = hwy)) + 
  geom_point(aes(color = factor(cyl))) +
  # specifying color to factor cyl
  facet_wrap(~year) + 
  # facet_wrap() compares year
  labs(x = "City MPG", # x-axis label
       y = "Highway MPG", # y-axis label 
       title = "City vs. Highway MPG (1999 vs 2008)")  # title for the graph


# It compares the effectiveness of cars on city and highways. It is better in 2008 compared to 1999 maybe because of advancement of technology. 
```

7) Come up with a chat by ploting two or more variables and tell me what it shows us. This can be a point plot, box, histogram, or another plot we covered. 
```{r}
ggplot(data, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = factor(cyl))) + 
  # specifying color to factor cyl
  labs(x = "Engine Displacement (L)", # x-axis label
       y = "Highway MPG", # y-axis label 
       title = "Highway MPG vs Engine Displacement") +  # title for the graph

  theme_bw()
#we can  observe that larger engines tend to have lower highway MPG, while smaller engines generally perform better on the highway in terms of fuel efficiency. The points are clustered to show this inverse relationship.

```

##Part 2:
Here we will use the data set 'gapminder' data set. This data set has basic country information.
```{r}
library(gapminder)
head(gapminder)
```

1) Explore the data and look at the variables listed, what are they?
```{r}
# Print the column names (variables) in the dataset
print(names(gapminder))  # Displays the names of all columns (variables)
# Display the first few rows of the dataset
head(gapminder)
# Summary statistics of the dataset
summary(gapminder)
```

2) Create a scatter plot of gdp v life expectancy for five (you pick) countries over a period of years. Plot these contries on the same chart. Show each country in a different color. Add a trend line (hint lm) for the set as a whole. Explain what you see.
```{r}

countries <- gapminder %>% # data set to filter and pipe
  filter(country %in% c("Japan", "India", "United States", "Australia", "Germany"))
    #Filter function filters countries from c vector provided 
#Creating the scatter plot with filtered countries
ggplot(countries, aes(x = gdpPercap, y = lifeExp, color = country)) + # aes() with x and y axes
  geom_point(alpha = 0.5) + # to see overlap used alpha (0.5) for semi-transparent points
  geom_smooth(method = "lm", se = FALSE, aes(group = 1, colour = "pink")) +  
  # using linear model line with color yellow and group = 1 for trend
  labs(x = "GDP per Capita", y = "Life Expectancy", title = "GDP vs Life Expectancy Across Five Countries") +
  # labels for x and y axes with title
  theme_bw()
#for good looking
#The trend line indicates a positive correlation between GDP and life expectancy, meaning that, in general, countries with higher GDP tend to have higher life expectancy. However, the slope of the trend line can vary, reflecting different economic growth patterns in these countries.
# There is a positive relationship involving GDP per capita and life expectation which indicates the country with higher gdp has extended life probability. As well, there is a substantial sum of flexibility in the plot which says other factors can be affecting the correlation like weather, lifestyle, diet, etc.
```

3) (Not graded) What was one thing you found interesting this week? What would you like changed? 
```{r}

```

