---
title: "DataMiningWeekTwo"
author: "Ajay Kommineni"
date: "Mon 31st,2025"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install and load packages

```{R, include=FALSE}
pkgs <- c("arules", "GGally", 
          "ggcorrplot", "hexbin", "palmerpenguins", "plotly", "seriation", "tidyverse")

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
library(arules)
library(GGally)
library(ggcorrplot)
library(hexbin)
library(palmerpenguins)
library(plotly)
library(seriation)
library(tidyverse)
```
# load the iris dataset

```{R}
library(tidyverse)
data(iris)
iris <- as_tibble(iris)
iris
```


```{r}
library(palmerpenguins)
data(package = 'palmerpenguins')
```


```{r}
head(penguins)
```
#Exercises
#1. Group the penguins by species, island or sex. What can you find out?

```{r}
# Load necessary libraries
library(palmerpenguins)  # Provides the penguin dataset
library(tidyverse)       # Includes dplyr and ggplot2 for data manipulation and visualization
```
# Grouping by species and summarizing key statistics
```{r}
penguins %>%                  
  group_by(species) %>%        
  # Grouping  the data by the 'species' column in the given data 
  summarise(                   
    count = n(),               
    #it  count the number of observations for each species recorded 
    avg_bill_length = mean(bill_length_mm, na.rm = TRUE),  
    # Calculates the  average bill length and ignoring Null values
    avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE),  
    # Calculates the  average bill depth and ignoring Null values
    avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE), 
    # Calculates the  average flipper length and ignoring Null values
    avg_body_mass = mean(body_mass_g, na.rm = TRUE)       
    # # Calculates the  average body of mass penguins  and ignoring Null values
  )

#findings
#1.The dataset contains three species: Adelie, Chinstrap, and Gentoo.
#2.Each species has a different number of recorded observations, indicating variations in their population representation in the dataset.
#3 Across all groups, body mass is positively correlated with flipper length—meaning larger penguins tend to have longer flippers and bills.
#The dataset highlights clear differences in bill size, flipper length, and body mass among penguin species. These differences may be related to habitat preferences, feeding behavior, and evolutionary adaptations

```

# Grouping by island and summarizing key statistics

```{r}
penguins %>%                  
  group_by(island) %>%        
  summarise(                  
    count = n(),               
    # Count the number of observations for each island
    avg_bill_length = mean(bill_length_mm, na.rm = TRUE),  
    avg_body_mass = mean(body_mass_g, na.rm = TRUE)      
  )
#Findings
#Some islands have a higher number of recorded penguins, suggesting species concentration differences across islands.
#Biscoe Island tends to have the highest number of penguins, indicating a favorable habitat for multiple species.
#This suggests that environmental conditions (food availability, temperature, etc.) could impact penguin growth and survival.
```

# Grouping by sex and summarizing key statistics

```{r}
penguins %>%                   
  group_by(sex) %>%          
  summarise(                 
    count = n(),              
    avg_body_mass = mean(body_mass_g, na.rm = TRUE)      
  )
#Findings
#the dataset shows that male penguins tend to have a higher body mass than females, a pattern common in many bird species. However, missing sex data could impact the accuracy of these findings, so further data cleaning or imputation may be necessary for precise analysis.

```

#Exercise 2
#Create histograms and boxplots for each continuous variable. Interpret the distributions.

#1. Bill Length (mm)
```{r}
library(ggplot2)
library(patchwork)
# Histogram: Shows frequency distribution of bill lengths
p1 <- ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  #binwidth = 1: Controls histogram bar width 
  #fill = species: Colors boxplots by species for easy comparison.
  labs(title = "Distribution of Bill Length", x = "Bill Length (mm)")

# Boxplot: Compares bill lengths across species
p2 <- ggplot(penguins, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Length by Species", y = "Bill Length (mm)")

p1 + p2

#interpretation
#Histogram:Bimodal distribution with peaks at ~35 mm and ~48 mm.Indicates two dominant groups:Smaller-billed penguins (Adelie) and Larger-billed penguins (Chinstrap/Gentoo).
#Boxplot:Adelie: Tightly clustered (median ~38 mm), minimal variability and Chinstrap/Gentoo: Overlap but Gentoo bills are slightly shorter on average.
```

#2Histogram & Boxplot for Bill Depth
```{r}
# Histogram
p3 <- ggplot(penguins, aes(x = bill_depth_mm)) +
  geom_histogram(binwidth = 0.5, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Bill Depth", x = "Bill Depth (mm)")

# Boxplot by species
p4 <- ggplot(penguins, aes(x = species, y = bill_depth_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Bill Depth by Species", y = "Bill Depth (mm)")

p3 + p4
#interpretation
#Histogram:Right-skewed with a peak at ~18 mm.Most penguins have medium-depth bills; few have very shallow/deep bills.
#Boxplot:Gentoo: Distinctly shallower bills (median ~15 mm), low variability.Adelie/Chinstrap: Similar depths but Adelies are slightly deeper.
```

#3. Flipper Length (mm)
#Histogram & Boxplot

```{r}
# Histogram
p5 <- ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 5, fill = "grey", color = "black") +
  labs(title = "Distribution of Flipper Length", x = "Flipper Length (mm)")

# Boxplot by species
p6 <- ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Flipper Length by Species", y = "Flipper Length (mm)")

p5 + p6
#interpertation
#Histogram:Trimodal distribution with peaks at ~190 mm, ~195 mm, and ~215 mm.Matches the three species: Adelie, Chinstrap, and Gentoo.
#Boxplot:Gentoo: Longest flippers (median ~217 mm), narrow range — efficient for deep diving.Adelie: Shortest flippers (median ~190 mm), high variability.
```

#4. Body Mass (g)
#Histogram & Boxplot

```{r}
# Histogram
p7 <- ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 100, fill = "blue", color = "black") +
  labs(title = "Distribution of Body Mass", x = "Body Mass (g)")

# Boxplot by species
p8 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(title = "Body Mass by Species", y = "Body Mass (g)")
p7 + p8
#interpertation
#Histogram:Bimodal distribution with peaks at ~3500 g (Adelie/Chinstrap) and ~5000 g (Gentoo).
#Boxplot:Gentoo: Heaviest (median ~5000 g), tightly clustered — consistent body size.Adelie/Chinstrap: Overlap, but Adelies show more variability.
```
#Exercise 3
#Create scatterplots and a scatterplot matrix. Can you identify correlations?
```{r}
# Load required packages
library(ggplot2)      # For creating plots
library(GGally)       # For scatterplot matrix (ggpairs)
library(palmerpenguins) # Contains the dataset

# Remove rows with missing values for clean analysis
penguins_clean <- na.omit(penguins)
```
#1 Bill Length vs. Bill Depth
```{r}
ggplot(penguins_clean, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2, alpha = 0.7) +  # Adjust point transparency and size
  labs(
    title = "Bill Length vs. Bill Depth by Species",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  ) +
  theme_minimal()
#Findings
#Gentoo	-0.53	Longer bills = shallower
#Adelie	+0.39	Longer bills = deeper
#Chinstrap	+0.30	Slight positive trend
```

#2. Flipper Length vs. Body Mass

```{r}
ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 2) +
  labs(
    title = "Flipper Length vs. Body Mass",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  )
#findings
#Strong Positive Correlation between Flipper Length vs. Body Mass
#Correlation (r): ~0.87 (all species)
#Longer flippers = heavier penguins. Flipper length scales with body size for efficient swimming.
```

#3. Scatterplot Matrix (All Variables)
```{r}
# Create a scatterplot matrix with species coloring
ggpairs(
  penguins_clean,
  columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g"),
  mapping = aes(color = species, alpha = 0.7),
  upper = list(continuous = wrap("cor", size = 3))  # Show correlation coefficients
) +
  labs(title = "Penguin Morphology Scatterplot Matrix") +
  theme(axis.text = element_text(size = 6))  # Adjust font size


```
#4. Create a reordered correlation matrix visualization. What does the visualizations show?

```{r}
library(tidyverse)
library(corrplot)

# Select numeric variables and compute correlations
cor_matrix <- penguins %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  na.omit() %>%
  cor()

# Reorder by hierarchical clustering
ord <- corrMatOrder(cor_matrix, order = "hclust")
reordered_cor_matrix <- cor_matrix[ord, ord]
```

```{r}
corrplot(
  reordered_cor_matrix,
  method = "color",         
  type = "upper",           # Show upper triangle
  order = "hclust",         # Cluster variables
  addCoef.col = "black",    # Add correlation coefficients
  tl.col = "darkblue",      # Variable label color
  tl.srt = 45,              # Rotate labels 45°
  title = "Reordered Correlation Matrix (Clustered)"
)
```











